---
title: "Ozgur's Data Analysis"
author: "Ali Nehme"
date: "`r format(Sys.time(), '%B %d, %Y %H:%M:%S', 'Canada', T)`"
output:
  html_document:
    df_print: tibble
    toc: true
    toc_float: 
          collapsed: false
          smooth_scroll: false
    number_sections: true
    theme: united
    #highlight: tango
    code_folding: show
    
  
---

```{r setup, include=F, cache=F}
knitr::opts_chunk$set(
	error = TRUE,
	message = FALSE,
	warning = FALSE,
	include = T
)

library(limma)
library(ggthemes)
library(ggrepel)
library(ade4)
library(factoextra)
library(Rtsne)
library(UpSetR)
library(EnhancedVolcano)
library(apeglm)
library(tidyverse)
library(enrichR)
library(DT)
#library(plyr)
library(tidyr)
library(tibble)
library(gridExtra)
library(grid)
library(clusterProfiler)
library(annotate)
library(org.Hs.eg.db)
library(msigdbr)
library(ggplot2)
library(RColorBrewer)
library(enrichplot)
library(reshape2)
library(pheatmap)
library(ggpubr)
library(ggbeeswarm)
library(ggthemes)
library(Biobase)
library(WGCNA)
library(DOSE)
source('E:/R directory/Scripts/GSEA_modified.R')
library(GSVA)
library(vsn)
source('E:/R directory/Scripts/geom_split_violin.R')
library(ggpubr)
#conflicted::conflict_prefer("select", "dplyr")

options(DT.options = list(extensions = c("Buttons" , "FixedColumns"),
                          autoWidth = TRUE , 
                           pageLength = 20,
                           searchHighlight = TRUE,
                           buttons = c('copy', 'txt', 'print'),
                           scrollX = F,
                          class = c('compact cell-border stripe hover'),
                          rownames = F, filter = 'top')
           )

```


# Data preparation

```{r Data preparation, echo=TRUE, message=FALSE}

#read expression matrix
q3data = read.table('CountMatrix_Q3.Norm.txt', header=T, sep='\t',row.names=1)
log2.data=log2(q3data)
scale.data = scale(log2(q3data))
vsn.data=normalizeVSN(q3data)

#read annotation file
pData = read.table("pData.txt", row.names=1, header =T)
pData= subset(pData, rownames(pData)%in%colnames(q3data))
for(p in c('Tumor_ratio_green', "Immune_ratio_red", "Stromal_ratio_yellow")){
  pData[,p] = as.numeric(pData[,p])
}

#make sure the samples order is consistent between the expression matrix and the samples annotation
if(!all.equal(colnames(q3data), rownames(pData))){print("Error: colnames(q3data) != rownames(pData)")}
stopifnot(all.equal(colnames(q3data), rownames(pData)))


HALLMARK_genesets = read.table("HALLMARK_genesets.txt", sep = "\t", header = T, row.names = 1)

table.include = c("Description", 'setSize' ,"enrichmentScore", "pvalue", "p.adjust")

```


# Sample visualization
## Data distribution
```{r echo=TRUE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE, , echo=TRUE, paged.print=FALSE}
data.list = c('log2.data', 'q3data','scale.data','vsn.data')
####Density plots
plots = lapply(data.list, function(x) 
ggplot(data.frame(melt(data.frame(get(x))), Group = rep(pData$LOX_relative_expression, each = nrow(q3data))),
       aes(x = value, group = variable, color = Group))+geom_density()+labs(title = x)+theme_few()
)
marrangeGrob(plots, ncol=1, nrow = length(data.list),
             newpage = F, gp = gpar(fontsize = 18,  face = "bold"))

#Violin plots
plots = lapply(data.list, function(x) 
  ggplot(data.frame(melt(data.frame(get(x))), Group = rep(pData$LOX_relative_expression, each = nrow(q3data))),
         aes(x = variable, y=value, fill = Group))+geom_violin()+geom_boxplot(width=0.1, fill = "grey")+
    labs(title = x)+theme_few())
marrangeGrob(plots, ncol=1, nrow = length(data.list),
             newpage = F, gp = gpar(fontsize = 18,  face = "bold"))


lapply(data.list, function(x) {meanSdPlot(as.matrix(get(x)))$gg + labs(title = x)})


```

VSN transformation seems to perform the best among the other transformation methods.

## PCA

```{r echo=TRUE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE, , echo=TRUE, paged.print=FALSE}

#create an expressionset object
if(all.equal(colnames(vsn.data), rownames(pData))){
  eset.vsn = ExpressionSet(assayData= as.matrix(vsn.data),phenoData=new("AnnotatedDataFrame",data=pData))
}

for(x in  c('vsn.data')) {

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)

df_pca <- prcomp(get(x))
df_out <- as.data.frame(df_pca$rotation)[,1:3]
df_out = cbind(df_out, pData)
df_out$repel = paste0(df_out$TMA, '_', df_out$Tumor_ratio_green,'%')

percentage <- round(df_pca$sdev / sum(df_pca$sdev) * 100, 2)
percentage <- paste0(colnames(df_out), " (", paste(as.character(percentage), " %", ")"))

variables = c('TMA', 'LOX_relative_expression', 
              'Tumor_ratio_green',
              "Histology")

plots = lapply(variables, function(p)
  if(is.factor(df_out[,p])){
    ggplot(df_out, aes(PC1, PC2))+
      labs(title=p, x=percentage[1], y = percentage[2], col=p)+
      geom_point(size=2, aes(col=factor(df_out[,p])))+
      scale_color_colorblind()+
      stat_ellipse(aes(col=df_out[,p]))+
      theme_minimal()+
      geom_text_repel(data=subset(df_out, rownames(df_out)%in%c('AOI22', 'AOI23', 'AOI24')),
                      aes(label=repel),
                      box.padding = unit(0.35, "lines"),
                      point.padding = unit(0.3, "lines"))
    
  }
  else if(is.numeric(df_out[,p]) | is.integer(df_out[,p])){
    ggplot(df_out, aes(PC1, PC2))+
      labs(title=p, x=percentage[1], y = percentage[2], col=p)+
      geom_point(size=2, aes(col=as.numeric(df_out[,p])))+
      scale_colour_gradient(low=colors[200], high = colors[1])+
      stat_ellipse(aes(col=df_out[,p]))+
      theme_minimal()+
      geom_text_repel(data=subset(df_out, rownames(df_out)%in%c('AOI22', 'AOI23', 'AOI24')),
                      aes(label=repel),
                      box.padding = unit(0.35, "lines"),
                      point.padding = unit(0.3, "lines"))
  }
)


print(marrangeGrob(plots, nrow=length(variables)/2, ncol=2, 
                   newpage = F, 
                   top = textGrob(paste0(x), gp = gpar(fontsize = 18,  face = "bold")))
)
}


```



There are 2 outlier samples that need to be removed. remove them before proceeding in the analysis.
There are 3 samples per condition. summarize each condition using the average of the 3 samples.

## collapsing

```{r echo=TRUE, fig.height=6, fig.width=12, message=FALSE, warning=FALSE, , echo=TRUE, paged.print=FALSE}

eseti = eset.vsn[,!colnames(eset.vsn)%in%c('AOI22', 'AOI23')]

collapsed.matrix  = t(collapseRows(t(exprs(eseti)),
                         rowGroup=pData(eseti)$TMA,
                         rowID=colnames(eseti),
                         method = "Average",
                         #method="function", methodFunction=psych::geometric.mean,
                         connectivityBasedCollapsing=F)$datETcollapsed)

#collapsed.matrix = scale(collapsed.matrix)

new.pData = pData(eseti)
new.pData = new.pData[!duplicated(new.pData$TMA),]; rownames(new.pData) = new.pData$TMA

if(all.equal(colnames(collapsed.matrix), rownames(new.pData))){
eset.collapsed = ExpressionSet(assayData= as.matrix(collapsed.matrix), phenoData=new("AnnotatedDataFrame",data=new.pData))}

vioplot::vioplot(collapsed.matrix)
affycoretools::plotPCA(eset.collapsed,
                        groups = as.numeric(eset.collapsed$LOX_relative_expression),
                        groupname = levels(eset.collapsed$LOX_relative_expression))

knitr::kable(summary(pData(eset.collapsed))) 

```



# Differential gene expression

```{r echo=TRUE, warning=FALSE, , message=FALSE, paged.print=FALSE}
rm(eset, eseti)

eseti = eset.collapsed[, eset.collapsed$LOX_relative_expression!="Intermediate"]
eseti$LOX_relative_expression = factor(eseti$LOX_relative_expression, levels = c("Low", "High"))

deg.table = data.frame()

eseti$LOX_relative_expression = relevel(eseti$LOX_relative_expression, ref = "Low")
design=model.matrix(~LOX_relative_expression, eseti)
colnames(design) = gsub("LOX_relative_expression", "", colnames(design))
knitr::kable(data.frame(design, X =" x ",pData(eseti)[, c('LOX_relative_expression', 'TMA', 'Histology')], check.names=F))
fit = lmFit(eseti, design)
fit = eBayes(fit)

deg=topTable(fit, coef="High", number=100, sort.by = "P", resort.by = "logFC") 
deg.table = rbind(deg.table, data.frame(Group = "High_vs_Low", deg[deg$P.Value<0.05,]))

deg.table$subGroup = factor(ifelse(deg.table$logFC>0, paste0(deg.table$Group,".up"),paste0(deg.table$Group,".down")))

cat("DEG: p<0.05")
summary(deg.table$subGroup[deg.table$P.Value<0.05])
cat("\n\n DEG: p.adjust<0.05")
summary(deg.table$subGroup[deg.table$adj.P.Val<0.05])
print(deg.table[deg.table$adj.P.Val<0.05,1:5])

DEGlist = split(rownames(deg.table),deg.table$subGroup)
DEGlist2 = split(rownames(deg.table),deg.table$Group)

write.table(deg.table, "Ozgur_DEGs.txt", sep = "\t", col.names = NA)

rm(design, fit, eseti, deg, fit2)

```


## Heatmaps
```{r echo=TRUE, fig.height=10, fig.width=12, message=FALSE, warning=FALSE, paged.print=FALSE}

print(summary(DEGlist)[,1, drop = F])

colors <- colorRampPalette(rev(brewer.pal(9, "RdBu")))(255)

eseti = eset.collapsed[, eset.collapsed$LOX_relative_expression!="Intermediate"]
eseti$LOX_relative_expression = factor(eseti$LOX_relative_expression, levels = c("Low", "High"))
annotation_colors = list(
  LOX_relative_expression = c(Low="gold", High = "darkred"))
  
for(i in names(DEGlist2)){
  
  cat(paste0("\n\n## ", 'DEG: ', i,'\n\n'))
  
  rm(yfg, ggdata, hm.data)
  yfg = DEGlist2[[i]]
  if(length(yfg>0)){
  
  #heatmap
  hm.data = data.frame(exprs(eseti))
  hm.data = hm.data %>% #log2 %>%
    subset(rownames(hm.data) %in% yfg)
  hm.data = t(scale(t(hm.data), center = T, scale =T))
  range <- max(abs(hm.data))
  annot_col = pData(eseti)[,c("LOX_relative_expression", 'Histology', "TMA")]
  sig.genes = rownames(deg.table)[deg.table$adj.P.Val<0.05]
  annot_row = tibble("adj.p<0.05" = factor(ifelse(rownames(hm.data)%in%sig.genes, 1,NA)), 
                     Gene = rownames(hm.data),
                     labels = factor(ifelse(rownames(hm.data)%in%sig.genes,rownames(hm.data), ""))) %>% 
    column_to_rownames('Gene')
  
  print(pheatmap(hm.data, main = paste0("DEGs: ", i), cluster_cols = T,
           clustering_distance_rows = "manhattan", clustering_distance_cols = "manhattan",
           col = colors, 
           annotation_col = annot_col,
           annotation_row = annot_row[,"adj.p<0.05",drop=F], labels_row = annot_row$labels,
           fontsize_row = 8,
           border_color = "white", 
           show_rownames = T, show_colnames = F, 
           breaks = c(seq(range.default(hm.data)[1], -0.01, length.out = 255/2),
                      seq(0.01, range.default(hm.data)[2], length.out = 255/2))
           ))
  
  }
}

```


## Overrepresentation analysis
```{r echo=TRUE, fig.height=10, fig.width=12, message=FALSE, warning=FALSE, paged.print=FALSE}

colors <- colorRampPalette(rev(brewer.pal(9, "Reds")))(100)

for(g in c('HALLMARK_genesets')){
  ck = compareCluster(geneCluster = DEGlist, fun = "enricher", TERM2GENE=get(g),
                          pAdjustMethod = "BH",
                          universe = rownames(eseti),
                          qvalueCutoff = 0.05
                         )
  print(dotplot(ck, showCategory = NULL)+
    labs(title=paste0(g, ': DEG sets'))+
      scale_color_continuous(low = colors[1], high = colors[75],
                               name = "p.adjust", guide=guide_colorbar(reverse=T)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)))
}

  
```


# GSEA: Hallmark of cancer
```{r echo=TRUE, fig.height=10, fig.width=14, message=FALSE, warning=FALSE, paged.print=FALSE}

eseti = eset.collapsed[, eset.collapsed$LOX_relative_expression!="Intermediate"]
eseti$LOX_relative_expression = factor(eseti$LOX_relative_expression, levels = c("Low", "High"))
eseti$LOX_relative_expression = relevel(eseti$LOX_relative_expression, ref = "Low")
design=model.matrix(~LOX_relative_expression, eseti)
colnames(design) = gsub("LOX_relative_expression", "", colnames(design))
fit = lmFit(eseti, design)
fit = eBayes(fit)
deg = topTable(fit, coef="High", number=Inf) #ADM vs Normal

#fgsea
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
colors = brewer.pal(8, "Set1")

genelist = deg$t
names(genelist) = rownames(deg)
genelist = genelist[order(genelist, decreasing = T)]
#genes <- rownames(deg)[deg$adj.P.Val<0.05]
for(gs in c('HALLMARK_genesets')){
gsea.res = GSEA(genelist, TERM2GENE = get(gs), pvalueCutoff = 0.25, verbose=FALSE)
print(gseaplot2(gsea.res, 
                geneSetID = rownames(gsea.res@result)[gsea.res@result$enrichmentScore>0],
                pvalue_table = TRUE, title = paste0(gs, ": High LOX"), subplots = 1:2,
                color = colors))
print(gseaplot2(gsea.res, 
                geneSetID = rownames(gsea.res@result)[gsea.res@result$enrichmentScore<0],
                pvalue_table = TRUE, title = paste0(gs, ": Low LOX"), subplots = 1:2,
                color = colors))
  }
  


  
```



# TF regulons: GSEA

```{r echo=TRUE, fig.height=10, fig.width=14, message=FALSE}

#geneset preparation
for(i in c(#'ChEA_2016',
           'ENCODE_and_ChEA_Consensus'#,
           #'ENCODE_TF_ChIP_seq_2015', 'TRANSFAC_and_JASPAR_PWMs', 'TRRUST_Transcription_Factors_2019'
           )){
  assign(i, read.gmt(paste0("Enrichr Genesets/", i, ".txt")))
}


#stats
eseti = eset.collapsed[, eset.collapsed$LOX_relative_expression!="Intermediate"]
eseti$LOX_relative_expression = factor(eseti$LOX_relative_expression, levels = c("Low", "High"))
eseti$LOX_relative_expression = relevel(eseti$LOX_relative_expression, ref = "Low")
design = model.matrix(~LOX_relative_expression, eseti)
colnames(design) = gsub("LOX_relative_expression", "", colnames(design))
fit = lmFit(eseti, design)
fit = eBayes(fit)
deg=topTable(fit, coef="High", number=Inf) #ADM vs Normal

genelist = deg$t
names(genelist) = rownames(deg)
genelist = genelist[order(genelist, decreasing = T)]

#GSEA
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
colors = brewer.pal(8, "Set1")

##TF regulons
for(gs in c(#'GTRD.TF.regulons', 'GTRD.miRNA.regulons', 'ChEA_2016',
            'ENCODE_and_ChEA_Consensus'#,
           #'ENCODE_TF_ChIP_seq_2015', 'TRANSFAC_and_JASPAR_PWMs', 'TRRUST_Transcription_Factors_2019'
           )){
gsea.res = GSEA(genelist, TERM2GENE = get(gs), pvalueCutoff = 0.25, verbose=FALSE)
gsea.df = gsea.res@result

if(nrow(gsea.df)!=0){
up.genesets = rownames(top_n(gsea.df[gsea.df$enrichmentScore>0, ], -10, p.adjust))
down.genesets = rownames(top_n(gsea.df[gsea.df$enrichmentScore<0, ], -10, p.adjust))

if(length(up.genesets)!=0){
print(gseaplot2(gsea.res, 
                geneSetID = up.genesets,
                pvalue_table = TRUE, title = paste0(gs, ": High LOX (top 10)"), subplots = 1:2,
                color = colors))}

if(length(down.genesets)!=0){
print(gseaplot2(gsea.res, 
                geneSetID = down.genesets,
                pvalue_table = TRUE, title = paste0(gs, ": Low LOX (top 10)"), subplots = 1:2,
                color = colors))}


print(cnetplot(gsea.res, foldChange = as.matrix(deg)[,'logFC']))

}
}


```


# Validation of top candidates using survival data

```{r echo=TRUE, fig.height=10, fig.width=12, message=FALSE}

library(RTCGA.clinical)
names(BRCA.clinical)[1:50]
clin.data <- survivalTCGA(BRCA.clinical,extract.cols=c("admin.disease_code", 
                                                       "patient.breast_carcinoma_progesterone_receptor_status"))
table(clin.data$patient.vital_status)

library(RTCGA.mRNA)
dim(BRCA.mRNA)


BRCA.mRNA.clin = BRCA.mRNA %>% 
                  as_tibble() %>% 
                  dplyr::select(bcr_patient_barcode, IRF8, E2F4, GABPA, MAX, E2F6) %>% 
                  mutate(bcr_patient_barcode = substr(bcr_patient_barcode, 1, 12)) %>% 
                  inner_join(clin.data, by="bcr_patient_barcode") %>%
                  mutate(progesterone_receptor_status = patient.breast_carcinoma_progesterone_receptor_status) %>% 
                  dplyr::select(-patient.breast_carcinoma_progesterone_receptor_status) %>%
                  mutate(IRF8_cat = ifelse(IRF8>=median(IRF8), "High", "Low"),
                         E2F4_cat = ifelse(E2F4>=median(E2F4), "High", "Low"),
                         GABPA_cat = ifelse(GABPA>=median(GABPA), "High", "Low"),
                         MAX_cat = ifelse(MAX>=median(MAX), "High", "Low"),
                         E2F6_cat = ifelse(E2F6>=median(E2F6), "High", "Low"))

ggdata = BRCA.mRNA.clin %>% 
  as_tibble() %>% 
  gather(gene, expression, IRF8, E2F4, GABPA, MAX, E2F6) %>%
  filter(!is.na(progesterone_receptor_status)) %>%

print(ggplot(ggdata, aes(progesterone_receptor_status, expression, fill=progesterone_receptor_status)) +
  geom_boxplot()+
  facet_wrap(~gene)
)

library(survival)
library(survminer)
coxph(Surv(times, patient.vital_status)~progesterone_receptor_status + IRF8 + E2F4 + GABPA + MAX + E2F6, 
      data=BRCA.mRNA.clin)

for(gene in c('IRF8', 'E2F4', 'GABPA', 'MAX', 'E2F6')){
  surv.formula = Surv(BRCA.mRNA.clin$times, BRCA.mRNA.clin$patient.vital_status)
  gene.exp = factor(as.data.frame(BRCA.mRNA.clin)[, paste0(gene, "_cat")])
  survdata = subset(BRCA.mRNA.clin, select = names(BRCA.mRNA.clin) %in% c(paste0(gene, "_cat"), 'times', 'patient.vital_status'))
sfit = survfit(surv.formula~gene.exp)
print(ggsurvplot(sfit, conf.int=F, pval=TRUE, data = survdata, legend.title = gene))
}
  
```

# Session info

```{r echo=TRUE, fig.height=10, fig.width=12, message=FALSE}

sessionInfo()
  
```